\documentclass{article}

\usepackage[a4paper,margin=2cm]{geometry}

\title{COMP30024 Teamwork Plan}
\date{\today}
\author{Lucas Fern (1080613),\\Hugo Lyons Keenan (1081696)}

\begin{document}
\maketitle
\section{Approach}
\begin{enumerate}
    \item Make a player class which contains the logic for our player. The player class will create its own representation of the board state which we will largely be able to steal from the project part A code.
    \item Implement methods to send moves and update representation based on actions returned from the referee.
    \item Create a basic implementation which sends random moves to test our interaction with the referee.
    \item Implement adversarial search algorithm. Consider using libraries with existing machine learning implementations.
\end{enumerate}

\section{Sharing Workload}
\begin{itemize}
    \item Collaborate on code with GitHub.
    \item Use GitHub statistics tracking to ensure we both complete approximately equal amounts of work.
    \item Divide code sections and modules as we discover new things that need to be implemented.
\end{itemize}

\section{Communication}
There were no issues with communication in part A, we will maintain open communication channels throughout the entire project and talk whenever we need to collaborate on ideas.

\section{Important Deadlines}
\subsection{Lucas}
\begin{itemize}
    \item Week 9-10, COMP30027 Assignment 2
    \item April 30th, MAST30025 Assignment 2
    \item Week 11-12, MAST30025 Assignment 3
    \item April 21st, SWEN30006 Assignment 1
    \item Week 10-12, SWEN30006 Assignment 2
\end{itemize}
\subsection{Hugo}
\begin{itemize}
    \item Week 9-10, COMP30027 Assignment 2
    \item April 30th, MAST30025 Assignment 2
    \item Week 11-12, MAST30025 Assignment 3
    \item 21st May, MUSI30021 Quiz + Performance
\end{itemize}

\end{document}